---
title: "Health Expenditure % GDP"
author: "David Andai"
date: "10/15/2022"
output: github_document
---

When trying to figure out how to allocate our financial resources, we more often than not asses our expenses so as to develop a budget. Some items on the budget list end up having a bigger share of the portion depending on our various priorities. 
When it comes to matters health at a National level, we use the Current Health Expenditure as a share of GDP to provide an indication on the level of resources channeled to health relative to other uses.

But what is GDP?
Gross domestic product (GDP) is the standard measure of the value added created through the production of goods and services in a country during a certain period. 
(IMAGE)

As GDP factors in the National outlook of the economy, health spending measures the final consumption of health care goods and services (i.e. current health expenditure). This includes personal health care (curative care, rehabilitative care, long-term care, ancillary services and medical goods) and collective services (prevention and public health services as well as health administration), but excluding spending on investments (OECD 2022).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rio)
library(tidyverse)
library(janitor)
library(ggplot2)
library(ggthemes)
library(dplyr)
```

Health Expenditure as percentage of GDP Data
```{r message=FALSE, warning=FALSE}
##Import and Cleaning Data Set
setwd('C:/Users/user/Documents/R/GDP/GDP')
he_gdp <- import('API_SH.XPD.CHEX.GD.ZS_DS2_en_csv_v2_4499032.csv')

#Data Cleaning
##reshape data frame 2009-2019 columns
he_gdp2 <- he_gdp[-c(5:53)]
names(he_gdp2) <- NULL
names(he_gdp2) <- he_gdp2[1,]
##remove first row and rename The Columns
he_gdp2 <- he_gdp2 [-c(1),]
he_gdp2 <- he_gdp2 %>%
  janitor::clean_names()
#delete empty columns
he_gdp2 <- he_gdp2 [ , !names(he_gdp2)%in%
                     c("country_code", "indicator_name", "indicator_code", "na", "x2020", "x2021")]

```

Health spending shows the importance of the health sector in the whole economy and indicates the societal priority which health is given measured in monetary terms (WHO). This analysis focuses how Kenya compares to its neighbours in the East Africa Community. The latest data entry is from 2019, while a look at a 10 year trend is assesd....... 

```{r}
#Obtain East Africa Countries
ea_gdp <- he_gdp2 [he_gdp2$country_name %in% c("Kenya", "Tanzania", "Uganda",
                "Rwanda", "Congo, Dem. Rep.", "Burundi", "South Sudan", "Ethiopia"),]
#cleaning data
ea_gdp_19 <- ea_gdp [ , c('country_name', 'x2019')]
##round to 2 decimal places
ea_gdp_19[,'x2019']=round(ea_gdp_19[,'x2019'],2)

##Data Visualization of HE $ of GDP in 2019
plot1 <- ggplot(ea_gdp_19,
       aes(x = reorder(country_name,-x2019), y= x2019, fill = reorder(country_name,-x2019)))+
  geom_bar(stat = 'identity',show.legend = FALSE)+
  scale_fill_manual(values = c("Kenya" = "#FF0000",
                               "Tanzania" = "#5cac94",
                               "Uganda" = "#5cac94",
                               "Rwanda"= "#5cac94", 
                               "Congo, Dem. Rep."= "#5cac94", 
                               "Burundi"= "#5cac94",
                               "South Sudan"= "#5cac94",
                               "Ethiopia"= "#5cac94"))+
  theme_few()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  geom_text(aes(label = paste(format(x2019, nsmall = 2), "%")), vjust = -0.2)+
  labs(title="Healthcare Expenditure as Percentage of GDP", 
        subtitle ="In 2019 in EastAfrican Countries")+
  labs(x = "",
       y = "")
```

```{r dev = "png"}
plot1
```

In 2019, health expenditure as a share of GDP for Kenya was 4.6 %. Between 2010 and 2019, health expenditure as a share of GDP in Kenya was decreasing on average by 2.44% each year, although before that, it grew from 4.6 % in 2000 to 6.1 % in 2010.

```{r message=FALSE, warning=FALSE}
#Trend of Healthcare Expenditure % of GDP in 10years

#Pivot data;  Wide-to-long
##pivot_data
ea_gdp2 <- ea_gdp%>%
  pivot_longer(
    cols = "x2009":"x2019",
    names_to = "year",
    values_to = "gdp"
  )

#Convert year column to numeric
years <- as.numeric(str_match(ea_gdp2$year,"[0-9]+"))

ea_gdp2 <- cbind(ea_gdp2, years)

#Data Visualization
myColours2 = c("#040c04", "#4d372c","#5cac94","#FF0000","#4d3ec0","#acc6d8",
               "#24a4d4", "#ca5cdd")
plot2 <- ea_gdp2%>%
  mutate(isKenya = (country_name == "Kenya"))%>%
  ggplot(aes(x=years, y=gdp, color=country_name))+
  geom_line( aes (linetype = isKenya), size =1, alpha = 0.6)+
  labs (title = "East Africa Countires Health Expenditure  (% of GDP)",
        y = "HE (% 0f GDP)",
        x = "",
        color = "Country")+
  scale_x_continuous(breaks = 2009:2019)+
  theme_few()+
  scale_linetype_manual(values = c ("dashed", "solid"), guide ="none")+
  scale_color_manual(values = myColours2)
```

```{r message=FALSE, warning=FALSE, dev='png'}
plot2
```
Over the last few years, there has been a strain on global economy following the Covid-19 virus and now the Ukraine-Russia war. Such constraints forces the governments to cut down on spending, including health spending.